---
title: Mini Proyecto de Análisis Exploratorio de Datos
author:
  - name: Nicolás Camañes Antolín
  - name: Rubén Castillo Carrasco
  - name: Erik González Soler
  - name: Jesús Martínez Leal

# document options
journal: notspecified
type: article

# front matter
simplesummary: |
  En este proyecto de Análisis Exploratorio de Datos se explora un conjunto de datos real con el fin de obtener valiosas perspectivas y conclusiones.
abstract: |
  
  A single paragraph of about 200 words maximum. For research articles, 
  abstracts should give a pertinent overview of the work. We strongly encourage
  authors to use the following style of structured abstracts, but without 
  headings: 1) Background: Place the question addressed in a broad context and
  highlight the purpose of the study; 2) Methods: Describe briefly the main
  methods or treatments applied; 3) Results: Summarize the article's main 
  findings; and 4) Conclusion: Indicate the main conclusions or interpretations. 
  The abstract should be an objective representation of the article, it must not 
  contain results which are not presented and substantiated in the main text and 
  should not exaggerate the main conclusions.
  
# back matter
keywords: |
  Análisis exploratorio de datos; Patrones en datos; Encuesta de población; Limpieza de datos
acknowledgement: |
  Es necesario mostrar nuestro agradecimiento a ValgrAI por el apoyo en el pago de la matrícula para el Máster en Ciencia de Datos (UV). Además, no nos podemos olvidar del profesorado que nos permitirá formarnos en este ámbito.
authorcontributions: |
  Aquí tenemos que poner más o menos lo que ha participado cada uno.
abbreviations:
  - short: AED
    long: Análisis Exploratorio de Datos
    
bibliography: mybibfile.bib
appendix: appendix.tex
endnotes: false
output: 
  rticles::mdpi_article:
    extra_dependencies: longtable
---

# Introducción.

El conjunto de datos que se utiliza para el análisis pertenece al Instituto Nacional de Estadística (INE). En concreto, pertenece a una encuesta realizada a la población en el año 2021 que tiene como propósito el proporcionar información detallada sobre personas, viviendas y edificios que no puede obtenerse a través de registros administrativos. En nuestro caso, hemos decidido hacer la exploración de una parte de esta, relacionada con cuestiones que se les hizo a los adultos (personas de 16 años o más) en las viviendas aleatoriamente seleccionadas. Esta está disponible  [aquí](https://www.ine.es/dyngs/INEbase/es/operacion.htm?c=Estadistica_C&cid=1254736177092&menu=resultados&idp=1254735572981#!tabs-1254736195790).

## Carga de librerías y datos necesarios para el análisis

```{r setup, cache = F, echo = F, message = F, warning = F, tidy = F}
# Configuración general de chunks
library(knitr)
options(width = 100)
knitr::opts_chunk$set(echo = T, message = F, error = F, warning = F, comment = NA, dpi = 100, tidy = T, cache.path = '.cache/', fig.path = './figure/', include = T)
```

Como es habitual, cargamos las librerías que necesitamos al inicio del documento.

```{r librerias, message = T, include = T, echo = T}
# Carga de librerías necesarias con pacman
library(pacman)
pacman::p_load(readr, stringr, tidyr, dplyr, readxl, ggplot2, forcats)
```

Pasamos ahora a hacer la carga del conjunto de datos, presentado en .txt en la carpeta *data* incluida en el repositorio del proyecto. Asimismo, cargamos el fichero .xlsx que nos muestra el *Diseño de registro y valores válidos de las variables*. Este fichero nos será especialmente útil puesto que permitirá una automatización para convertir nuestra *Raw data* a *Technically correct data*, teniendo cada variable su tipo correcto.

```{r lectura_dataset}
ruta_txt <- "./data/md_ECEPOVadultos_2021.txt"
lines <- readLines(ruta_txt)

set.seed(33)
lines <- sample(lines, 40000) # si se desea carga rápida (quitar luego)
```

```{r lectura_diseno}
ruta_excel <- "data/dr_ECEPOVadultos_2021.xlsx"
excel_info <- read_excel(ruta_excel, sheet = "Diseño", range = "A2:H50", col_names = TRUE)

# Almacenamiento de distintas hojas del excel para optimizar el código
Tablas1 <- read_excel(ruta_excel, sheet = "Tablas1", range = "A4:C63", col_names = TRUE)
Tablas2 <- read_excel(ruta_excel, sheet = "Tablas2", range = "A4:C112", col_names = TRUE)
Tablas3 <- read_excel(ruta_excel, sheet = "Tablas3", range = "A4:C48", col_names = TRUE)
```

```{r extraccion_dataset}
# Función para sacar las subcadenas en cada línea del .txt

extract_substrings <- function(line, start_position, length) {
  substrings <- character(length(start_position))
  
  for (i in seq_along(start_position)) {
    start <- start_position[i]
    end <- start + length[i] - 1
    substrings[i] <- substr(line, start, end)
  }
  
  return(substrings)
}

# Guardaremos las substrings en una matriz por eficiencia computacional

num_lines <- length(lines)
num_variables <- length(excel_info$Variable)
substring_matrix <- matrix("", nrow = num_lines, ncol = num_variables)

# Extraer las subcadenas para todas las líneas y guardarlas en las filas de la matriz

for (i in 1:num_lines) {
  substring_matrix[i, ] <- extract_substrings(lines[i], excel_info$Posición, excel_info$Longitud)
}

# Conversión al dataframe desde la matriz

df <- as.data.frame(substring_matrix)
colnames(df) <- excel_info$Variable

```

A pesar de que fue de gran utilidad el fichero que ofrecía INE para la transformación de las variables a un tipo correcto, aparecieron algunos fallos que tuvimos que solventar de manera manual. Determinadas variables no fueron bien codificadas por parte de INE, de tal forma que hubo que modificar ligeramente el fichero .xlsx. Concretamente, fue en algunos *diccionarios* de variables en los que no terminaba de estar correctamente bien determinado el *código* (lo que sería el level en un factor). El error presentado (en algunas variables) consistía, básicamente, en que aparecía "01" en lugar de "1" (para otros números similar), valor recogido en las observaciones.

```{r conversion_dataset}
# Cambio manual por datos mal introducidos "2 " --> "2" para la posterior lectura con el diccionario

manual_change <- c("ESTUDIOS", "CAMPO", "NDESPLA", "MTRANSPOR_1", "MTRANSPOR_2")

df <- df %>%
  mutate(across(all_of(manual_change), ~ str_remove_all(., " ")))

# Conversión de las variables correspondientes a tipo numérico según indica INE

dic_tipo <- excel_info$Tipo

df <- df %>%
  mutate(across(which(dic_tipo == "N"), as.numeric)) %>%
  mutate()

# Transformación a factores de las variables pertinentes usando el fichero que ofrece INE

excel_dic <- excel_info %>%
  filter(!is.na(`Diccionario de la variable`))

dic_vars <- excel_dic$Variable
dic_var <- excel_dic$`Diccionario de la variable`
dic_table <- excel_dic$`Diccionario ubicado en la hoja…`

## bucle donde se irán convirtiendo a factor de manera iterativa aquellas variables que nos indique el fichero del INE

for (i in seq_along(dic_vars)) {
  
  dataframe_name <- dic_table[i]  
  df_table <- get(dataframe_name)
  positions <- which(df_table == dic_var[i], arr.ind = TRUE) # obtener posición donde se encuentra la palabra del diccionario 
  
  if (length(positions) > 0) {
    
    fila_start <- positions[1]
    columna_start <- positions[2]
    
    fila_end <- min(which(is.na(df_table[(fila_start + 2):nrow(df_table), columna_start]), arr.ind = TRUE)[1]) + fila_start

    levels <- df_table[(fila_start + 2):fila_end, columna_start]
    labels <- df_table[(fila_start + 2):fila_end, columna_start + 1]
    
    # conversión a tipo vector para aplicar debajo factor()
    levels <- unlist(levels[[1]])
    labels <- unlist(labels[[1]])
  }
  
  df <- df %>%
      mutate_at(vars(dic_vars[i]), ~factor(., levels = levels, labels = labels)) 
}
```

```{r}
df %>%
  select(MTRANSPOR_1) %>%
  filter(MTRANSPOR_1 == "Otros - andando")
```

# Ideas de Jesús

Univariante: distribución de provincia, estado civil, campos de estdios, situación laboral, medio de transporte usado para trabajo, dispone teléfono móvil, hijos


Journal options are in Table \ref{tab:mdpinames}. The `status` variable should 
generally not be changed by authors. The `type` variable describes 
the type of of submission and defaults to `article` but can be replaced with any of the ones in Table \ref{tab:mdpitype}

```{r mdpitype, echo = FALSE}
type <- c("abstract, addendum, article, book, bookreview, briefreport, 
casereport, comment, commentary, communication, conferenceproceedings, 
correction, conferencereport, entry, expressionofconcern, extendedabstract, 
datadescriptor, editorial, essay, erratum, hypothesis, interestingimage, 
obituary, opinion, projectreport, reply, retraction, review, perspective, 
protocol, shortnote, studyprotocol, systematicreview, supfile, 
technicalnote, viewpoint, guidelines, registeredreport, tutorial")
knitr::kable(rticles::string_to_table(type, 3), 
             col.names = NULL, format = "latex", booktab = TRUE, 
             caption = "MDPI article types.")
```



\startlandscape
```{r, mdpinames, echo=FALSE, results='asis'}
mdpi_journals <- "acoustics, actuators, addictions, admsci, adolescents, aerobiology, aerospace, agriculture, agriengineering, agrochemicals, agronomy, ai, air, algorithms, allergies, alloys, analytica, analytics, anatomia, animals, antibiotics, antibodies, antioxidants, applbiosci, appliedchem, appliedmath, applmech, applmicrobiol, applnano, applsci, aquacj, architecture, arm, arthropoda, arts, asc, asi, astronomy, atmosphere, atoms, audiolres, automation, axioms, bacteria, batteries, bdcc, behavsci, beverages, biochem, bioengineering, biologics, biology, biomass, biomechanics, biomed, biomedicines, biomedinformatics, biomimetics, biomolecules, biophysica, biosensors, biotech, birds, bloods, blsf, brainsci, breath, buildings, businesses, cancers, carbon, cardiogenetics, catalysts, cells, ceramics, challenges, chemengineering, chemistry, chemosensors, chemproc, children, chips, cimb, civileng, cleantechnol, climate, clinpract, clockssleep, cmd, coasts, coatings, colloids, colorants, commodities, compounds, computation, computers, condensedmatter, conservation, constrmater, cosmetics, covid, crops, cryptography, crystals, csmf, ctn, curroncol, cyber, dairy, data, ddc, dentistry, dermato, dermatopathology, designs, devices, diabetology, diagnostics, dietetics, digital, disabilities, diseases, diversity, dna, drones, dynamics, earth, ebj, ecologies, econometrics, economies, education, ejihpe, electricity, electrochem, electronicmat, electronics, encyclopedia, endocrines, energies, eng, engproc, entomology, entropy, environments, environsciproc, epidemiologia, epigenomes, est, fermentation, fibers, fintech, fire, fishes, fluids, foods, forecasting, forensicsci, forests, foundations, fractalfract, fuels, future, futureinternet, futurepharmacol, futurephys, futuretransp, galaxies, games, gases, gastroent, gastrointestdisord, gels, genealogy, genes, geographies, geohazards, geomatics, geosciences, geotechnics, geriatrics, grasses, gucdd, hazardousmatters, healthcare, hearts, hemato, hematolrep, heritage, higheredu, highthroughput, histories, horticulturae, hospitals, humanities, humans, hydrobiology, hydrogen, hydrology, hygiene, idr, ijerph, ijfs, ijgi, ijms, ijns, ijpb, ijtm, ijtpp, ime, immuno, informatics, information, infrastructures, inorganics, insects, instruments, inventions, iot, j, jal, jcdd, jcm, jcp, jcs, jcto, jdb, jeta, jfb, jfmk, jimaging, jintelligence, jlpea, jmmp, jmp, jmse, jne, jnt, jof, joitmc, jor, journalmedia, jox, jpm, jrfm, jsan, jtaer, jvd, jzbg, kidneydial, kinasesphosphatases, knowledge, land, languages, laws, life, liquids, literature, livers, logics, logistics, lubricants, lymphatics, machines, macromol, magnetism, magnetochemistry, make, marinedrugs, materials, materproc, mathematics, mca, measurements, medicina, medicines, medsci, membranes, merits, metabolites, metals, meteorology, methane, metrology, micro, microarrays, microbiolres, micromachines, microorganisms, microplastics, minerals, mining, modelling, molbank, molecules, mps, msf, mti, muscles, nanoenergyadv, nanomanufacturing, nanomaterials, ncrna, ndt, network, neuroglia, neurolint, neurosci, nitrogen, notspecified, nri, nursrep, nutraceuticals, nutrients, obesities, oceans, ohbm, onco, oncopathology, optics, oral, organics, organoids, osteology, oxygen, parasites, parasitologia, particles, pathogens, pathophysiology, pediatrrep, pharmaceuticals, pharmaceutics, pharmacoepidemiology, pharmacy, philosophies, photochem, photonics, phycology, physchem, physics, physiologia, plants, plasma, platforms, pollutants, polymers, polysaccharides, poultry, powders, preprints, proceedings, processes, prosthesis, proteomes, psf, psych, psychiatryint, psychoactives, publications, quantumrep, quaternary, qubs, radiation, reactions, receptors, recycling, regeneration, religions, remotesensing, reports, reprodmed, resources, rheumato, risks, robotics, ruminants, safety, sci, scipharm, sclerosis, seeds, sensors, separations, sexes, signals, sinusitis, skins, smartcities, sna, societies, socsci, software, soilsystems, solar, solids, spectroscj, sports, standards, stats, std, stresses, surfaces, surgeries, suschem, sustainability, symmetry, synbio, systems, targets, taxonomy, technologies, telecom, test, textiles, thalassrep, thermo, tomography, tourismhosp, toxics, toxins, transplantology, transportation, traumacare, traumas, tropicalmed, universe, urbansci, uro, vaccines, vehicles, venereology, vetsci, vibration, virtualworlds, viruses, vision, waste, water, wem, wevj, wind, women, world, youth, zoonoticdis"

knitr::kable(rticles::string_to_table(mdpi_journals, 8),
             col.names = NULL,
             format = "latex",
             booktabs = TRUE, 
             caption = "MDPI journal names.",
             longtable = TRUE)

```
\finishlandscape




# Introduction

The introduction should briefly place the study in a broad context and highlight
why it is important. It should define the purpose of the work and its
significance. The current state of the research field should be reviewed
carefully and key publications cited. Please highlight controversial and
diverging hypotheses when necessary. Finally, briefly mention the main aim of
the work and highlight the principal conclusions. As far as possible, please
keep the introduction comprehensible to scientists outside your particular
field of research. Citing a journal paper [@bertrand-krajewski_distribution_1998;
@leutnant_stormwater_2016]. And now citing a book reference @gujer_systems_2008.
Some MDPI journals use Chicago and others use APA, this template should choose
the correct citation format for you once you specify the journal in the YAML
header.

To use endnotes, change `endnotes: true` in the YAML header, then use
`\endnote{This is an endnote.}`.


# Materials and Methods

Materials and Methods should be described with sufficient details to allow 
others to replicate and build on published results. Please note that publication 
of your manuscript implicates that you must make all materials, data, computer 
code, and protocols associated with the publication available to readers. Please
disclose at the submission stage any restrictions on the availability of 
materials or information. New methods and protocols should be described in 
detail while well-established methods can be briefly described and appropriately
cited.

Research manuscripts reporting large datasets that are deposited in a publicly 
available database should specify where the data have been deposited and provide
the relevant accession numbers. If the accession numbers have not yet been
obtained at the time of submission, please state that they will be provided
during review. They must be provided prior to publication.

Interventionary studies involving animals or humans, and other studies require
ethical approval must list the authority that provided approval and the
corresponding ethical approval code.

# Results

This section may be divided by subheadings. It should provide a concise and
precise description of the experimental results, their interpretation as well
as the experimental conclusions that can be drawn.

## Subsection Heading Here

Subsection text here.

### Subsubsection Heading Here

Bulleted lists look like this:

* First bullet
* Second bullet
* Third bullet

Numbered lists can be added as follows:

1. First item
2. Second item
3. Third item

The text continues here.

## Figures, Tables and Schemes

All figures and tables should be cited in the main text as Figure \ref{fig:fig1}, 
\ref{tab:tab1}, etc. To get cross-reference to figure generated by R chunks 
include the `\\label{}` tag in the `fig.cap` attribute of the R chunk: 
`fig.cap = "Fancy Caption\\label{fig:plot}"`.

```{r fig1, echo=FALSE, fig.width=3, fig.cap="A figure added with a code chunk.\\label{fig:fig1}"}
x = rnorm(10)
y = rnorm(10)
plot(x, y)
```


When making tables using `kable`, it is suggested to use
the `format="latex"` and `tabl.envir="table"` arguments
to ensure table numbering and compatibility with the mdpi
document class.

```{r tab1, echo=FALSE}
knitr::kable(mtcars[1:5, 1:3], format = "latex", 
             booktabs = TRUE, 
             caption = "This is a table caption. Tables should be placed in the 
             main text near to the first time they are~cited.", 
             align = 'ccc', centering = FALSE,
             table.envir = "table", position = "H")
```



For a very wide table, landscape layouts are allowed.


\startlandscape

```{r tab2, echo=FALSE}
df <- data.frame(`Title 1` = c("Entry 1", "Entry 2"),
                 `Title 2` = c("Data", "Data"),
                 `Title 3` = c("Data", "Data"),
                 `Title 4` = c("This cell has some longer content that runs over
                               two lines",
                               "Data"))
knitr::kable(df, format = "latex", 
             booktabs = TRUE, 
             caption = "This is a very wide table", 
             align = 'ccc', centering = FALSE,
             table.envir = "table", position = "H")
```

\finishlandscape

## Formatting of Mathematical Components

This is an example of an equation:

$$
a = 1.
$$

If you want numbered equations use Latex and wrap in the equation environment:

\begin{equation}
a = 1,
\end{equation}

the text following an equation need not be a new paragraph. Please punctuate 
equations as regular text.

```{comment}
If the documentclass option "submit" is chosen, please insert a blank line before and after any math environment (equation and eqnarray environments). This ensures correct linenumbering. The blank line should be removed when the documentclass option is changed to "accept" because the text following an equation should not be a new paragraph.
```

This is the example 2 of equation:

\begin{adjustwidth}{-\extralength}{0cm}
\begin{equation}
a = b + c + d + e + f + g + h + i + j + k + l + m + n + o + p + q + r + s + t + 
u + v + w + x + y + z
\end{equation}
\end{adjustwidth}

Theorem-type environments (including propositions, lemmas, corollaries etc.) 
can be formatted as follows:

Example of a theorem:

::: {.Theorem latex=true}
Example text of a theorem
:::

The text continues here. Proofs must be formatted as follows:

Example of a proof:

::: {.proof latex="[Proof of Theorem1]"}
Text of the proof. Note that the phrase ``of Theorem 1'' is optional if it is 
clear which theorem is being referred to.
:::

The text continues here.

# Discussion

Authors should discuss the results and how they can be interpreted in 
perspective of previous studies and of the working hypotheses. The findings and 
their implications should be discussed in the broadest context possible. Future 
research directions may also be highlighted.

# Conclusion

This section is not mandatory, but can be added to the manuscript if the
discussion is unusually long or complex.

# Patents

This section is not mandatory, but may be added if there are patents resulting
from the work reported in this manuscript.
